\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-23 metatheory-condensed.tex
Change into p and var Add one sentence explain substitution.

$$
\Rule[name=Subst]
{\goodType{\Delta}{T}{j} 
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodType{\Gamma}{T[\gamma]}{j}},
\Rule[]
{\goodTerm{\Delta}{t}{T}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodTerm{\Gamma}{t[\gamma]}{T[\gamma]}}
\quad
\Rule[name=Dbj Indices]
{\goodSub{\Gamma}{\gamma}{(\Delta, A)}}
{\goodSub{\Gamma}{\pi_1 \gamma}{\Delta}},
\Rule[]
{\goodSub{\Gamma}{\gamma}{(\Delta, A)}}
{\goodSub{\Gamma}{\pi_2 \gamma}{A[\pi_1 \gamma]}}
$$

% Explicit substitution
Instead of meta-level substitution, we use \textbf{explicit substitution}, i.e., substitution itself is a expression in the program and will be done in the runtime. Thus substitution has its own judgement $\goodSub{\Gamma}{\gamma}{\Delta}$. We can apply substitution to types and terms\footnote{The type judgement is the form $\goodType{\Gamma}{T}{}$. The term judgement is the form $\goodTerm{\Gamma}{t}{T}$} in MLTT via \ruleref{Subst}. We also have an identity substitution $\goodSub{\Gamma}{"id"}{\Gamma}$.
% Debruijn Indices
Contrary to named variables, \textbf{debruijn indices} refer to the variable in the context like operating a stack. For example, we can get the very first variable in the context $\Gamma, x : T \vdash x : T$ by $\goodTerm{\Gamma, T}{\pi_2 "id"}{T[\pi_1"id"]}$. Similarly, to get the second and above variables, we only need to apply a sequence of $\pi_1$s.

@@

% linkage, signature and compilation and seal
$$
\Rule[name=Ept Sig]
{}
{\goodSig{\Gamma}{\nu\cdot}{0}}\quad
\Rule[name=Lnkg Type/Compile]
{\goodSig{\Gamma}{\sigma}{n}}
{\goodType{\Gamma}{\cL \sigma}{i}
\quad \goodType{\Gamma}{\cC \sigma}{}}
\quad
\Rule[name=Lnkg Compile]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
}
{
  \goodTerm{\Gamma}{\cCt o}{\cC \sigma}
}
$$
$$
\Rule[name=Sig Add]
{\goodSig{\Gamma}{\sigma}{n} 
 \quad \goodType{\Gamma}{A}{}
 \quad \goodSeal{\Gamma}{f}{\sigma}{A}
 \quad \goodType{\Gamma, A}{T}{i}}
{\goodSig{\Gamma}{(\nu^+ \ \sigma \ \{f\} \ T)}{n+1}}
\quad
\Biggl( \boxed{\goodSeal{\Gamma}{s}{\sigma}{A}} 
:= \boxed{\goodTerm{\Gamma, \cC\sigma}{s}{A[\pi_1]}} \Biggr)
$$
$$
\Rule[name=Ept Lnkg]
{}
{\goodTerm{\Gamma}{\mu\cdot}{\cL \nu\cdot}}
\quad
\Rule[name=Lnkg Add]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
\quad \goodType{\Gamma}{A}{}
\quad  {\goodSeal{\Gamma}{f}{\sigma}{A}}
 \quad \boxed{\goodTerm{\Gamma, A}{t}{T}}
}
{\goodTerm{\Gamma}{(\mu^+ \ o \ \{f\} \ t)}{\cL(\nu^+ \ \sigma \ \{f\} \ T)}}
$$
We start with signature (in a brand new judgement $\goodSig{\Gamma}{\sigma}{n}$). A signature is a list of types contained by a \textit{linkage}\footnote{Following the convention by \citet{zm2017}, we name the meta-theoretical correspondents of \textit{family} as \textit{linkage}.}, tracking the length as superscript.   We will use $\{a : T_1, b : T_2, ..\}$ as a shorthand for signature with type $T_1, T_2, ..$ for readability.

@@

  \quad \goodTerm{\Gamma, \pi^j_2\tau[(id, a)]}{b}{El~("W"~T)[\pi_1]}}

@@

{\goodType{\Gamma}{"RecSig1"~A~B~R \equiv \Pi A\; \Pi (\Pi B (R[\pi_1\circ\pi_1]))\; R[\pi_1\circ\pi_1]}{}}

@@

("app"~(("app"~("prjR"~j~h))[("id", a)]))[("id", "lam" ("Wrec"~h[\pi_1]~b))]

@@

\begin{align*}
  \text{We first define a new type }& \goodType{\Gamma}{\Sigr{\Gamma}{n}}{}  \text{ when } \goodCtx{\Gamma}{}  \text{ is well-formed and } n \in \mathbb{N} \\
  \denotesT{\goodSig{\Gamma}{\_}{n}},\ &({\Sigr{\Gamma}{n}}), \text{ and } \ \denotesT{\cC} \text{ are mutually recursive, } \\
  & \text{defined by induction on the signature length} \\  
  \denotesT{\goodSig{\Gamma}{\_}{n}} &= (\goodTerm{\Gamma}{\_}{\Sigr{\Gamma}{n}}) \text{, and thus } \denotesT{\goodSig{\Gamma}{\sigma}{n}} \text{defined} \iff \goodTerm{\Gamma}{\denotesT{\sigma}}{\Sigr{\Gamma}{n}} \\ 
  {\Sigr{\Gamma}{(n+1)}} &= 
    \Sigma~(\Sigr{\Gamma}{n})
          ~(\Sigma~\cU
                  ~(\Sigma~(\Pi(\cC (\pi_2[\pi_1]))(El~\pi_2[\pi_1]))~\cU)) \\
  \Sigr{\Gamma}{0} &= \top \\
  \denotesT{\goodType{\Gamma}{\cC \sigma}{}}, \denotesT{\goodType{\Gamma}{\cL\sigma}{}} &\text{ defines upon } \denotesT{\sigma} 
  \text{ and inductively on the signature length} \\
  \denotesT{\goodType{\Gamma}{\cC \sigma}{}} &= \denotesT{\goodType{\Gamma}{\cL\sigma}{}} = \top \quad
      \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{0}} \\ 
  \denotesT{\goodType{\Gamma}{\cC \sigma}{}} &= 
    \Sigma~\denotesT{\cC ("pjl"~\sigma)}~(El \ ("app"~("pjr"^3~\sigma))[(\pi_1, "app"~("pjl"~("pjr"^2~t)))]) \\
      &\text{given } \denotesT{\goodSig{\Gamma}{\sigma}{n+1}} \\
  \denotesT{\goodType{\Gamma}{\cL\sigma}{}} &=
  \denotesT{\cL("pjl"~\sigma)} \times \Pi(El~("pjl"~("pjr"~\sigma)))(El~("app"~("pjr"^3~\sigma))) \\
  \denotesT{\goodSig{\Gamma}{\nu^+~\sigma~f~T}{n+1}} &= (\sigma, "c"~A, "lam"~f, "lam"~("c"~T)) \\ 
  \denotesT{\goodTerm{\Gamma}{\mu^+~m~t}{\cL(\nu^+~\sigma~T)}} &= (m, "lam"~t) \\ 
  \denotesT{\goodTerm{\Gamma}{\cCt m}{\cC\sigma}} & \text{ is defined upon } \denotesT{m} \text{ and inductively on the signature length} \\ 
  \denotesT{\goodTerm{\Gamma}{\cCt m}{\cC\sigma}} &= 
  (\denotesT{\cCt o}, t["sf"~f][("id", \denotesT{\cCt o})]) \\
  & \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{n+1}}, \text{ where } o = "pjl"~m, t = "app"~("pjr"~m), f = p_f\nu~\sigma \\ 
  \denotesT{\goodTerm{\Gamma}{\cCt m}{\cC \sigma}} &= () \quad \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{0}}
\end{align*}
\begin{align*}
  \denotesT{\goodInh{\Gamma}{\_}{\sigma_1}{\sigma_2}} &= \goodTerm{\Gamma, \cL\sigma_1}{\_}{\cL\sigma_2[\pi_1]} \\ \text{ and thus } \denotesT{\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}} &\iff  \goodTerm{\Gamma, \cL\sigma_1}{h}{\cL\sigma_2[\pi_1]} \\
  \denotesT{\goodInh{\Gamma}{"inhinh"~h~T~\uparrow^s}{(\nu^+ \  \sigma_1 \  T)}{(\nu^+ \  \sigma_2 \  T[(\pi_1, \uparrow^s)])}} &= \mu^+~(h[(\pi_1,\pi_1\mu~\pi_2)])\\ &\quad \quad \{f_2[\pi_1]\}~(p_2\mu'~\pi_2)[(\pi_1, \uparrow^s[\pi_1^{\uparrow}])] \\
\end{align*}

>>>>

$$
\Rule[name=Subst]
{\goodType{\Delta}{T}{j} 
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodType{\Gamma}{T[\gamma]}{j}},
\Rule[]
{\goodTerm{\Delta}{t}{T}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodTerm{\Gamma}{t[\gamma]}{T[\gamma]}}
\quad
\Rule[name=Dbj Indices]
{}
{\goodSub{\Gamma, A}{p^1 \gamma}{\Gamma}},
\Rule[]
{}
{\goodSub{\Gamma, A}{"var"_0}{A[p^1]}}
$$

% Explicit substitution
Instead of meta-level substitution, we use \textbf{explicit substitution}, i.e., substitution itself is a expression in the program and will be done in the runtime. Thus substitution has its own judgement $\goodSub{\Gamma}{\gamma}{\Delta}$. We can apply substitution to types and terms\footnote{The type judgement is the form $\goodType{\Gamma}{T}{}$. The term judgement is the form $\goodTerm{\Gamma}{t}{T}$} in MLTT via \ruleref{Subst}.  Note how substitution will change the context---we suggest intuitvely understanding context as holes of the term/types and substitution as filling those holes, where the substitution itself has holes. After filling the holes, then the new context is about the holes of the substitution. We have an identity substitution $\goodSub{\Gamma}{"id"}{\Gamma}$.
% Debruijn Indices
Contrary to named variables, \textbf{debruijn indices} refer to the variable in the context like operating a stack. We will use "var"$_i$ indicating the $i$-th variable (from right to left, $i$ starts with 0) in the context.  On the other hand, we can also get the rest part of the stack, or ``subpart of "id"'' using $\goodSub{\Gamma, A}{p^1 \gamma}{\Gamma}$.  For example, we can get the very first variable in the context $\Gamma, x : T \vdash x : T$ by $\goodTerm{\Gamma, T}{var_0}{T[p^1]}$. Note here $T$ is a type in $\Gamma$, but now shifted into $\goodType{\Gamma, T}{T[p^1]}{}$ using $p^1$. Similarly, to get the second and prior variables, we use "var"$_i$ and $p^{i+1}$ correspondingly.

@@

% linkage, signature and compilation and seal
$$
\Rule[name=Ept Sig]
{}
{\goodSig{\Gamma}{\nu\cdot}{0}}\quad
\Rule[name=Lnkg Type/Compile]
{\goodSig{\Gamma}{\sigma}{n}}
{\goodType{\Gamma}{\cL \sigma}{i}
\quad \goodType{\Gamma}{\cC \sigma}{}}
\quad
\Rule[name=Lnkg Compile]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
}
{
  \goodTerm{\Gamma}{\cCt o}{\cC \sigma}
}
$$
$$
\Rule[name=Sig Add]
{\goodSig{\Gamma}{\sigma}{n} 
 \quad \goodType{\Gamma}{A}{}
 \quad \goodSeal{\Gamma}{f}{\sigma}{A}
 \quad \goodType{\Gamma, A}{T}{i}}
{\goodSig{\Gamma}{(\nu^+ \ \sigma \ \{f\} \ T)}{n+1}}
\quad
\Biggl( \boxed{\goodSeal{\Gamma}{s}{\sigma}{A}} 
:= \boxed{\goodTerm{\Gamma, \cC\sigma}{s}{A[p^1]}} \Biggr)
$$
$$
\Rule[name=Ept Lnkg]
{}
{\goodTerm{\Gamma}{\mu\cdot}{\cL \nu\cdot}}
\quad
\Rule[name=Lnkg Add]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
\quad \goodType{\Gamma}{A}{}
\quad  {\goodSeal{\Gamma}{f}{\sigma}{A}}
 \quad \boxed{\goodTerm{\Gamma, A}{t}{T}}
}
{\goodTerm{\Gamma}{(\mu^+ \ o \ \{f\} \ t)}{\cL(\nu^+ \ \sigma \ \{f\} \ T)}}
$$
We start with signature (in a brand new judgement $\goodSig{\Gamma}{\sigma}{n}$). A signature is a list of types contained by a \textit{linkage}\footnote{Following the convention by \citet{zm2017}, we name the meta-theoretical correspondents of \textit{family} as \textit{linkage}.}, tracking the length as superscript.   We will use $\{a : T_1, b : T_2, ..\}$ as a shorthand for signature with type $T_1, T_2, ..$ for readability.

@@

  \quad \goodTerm{\Gamma, \pi^j_2\tau[(id, a)]}{b}{El~("W"~T)[p^1]}}

@@

{\goodType{\Gamma}{"RecSig1"~A~B~R \equiv \Pi A\; \Pi (\Pi B (R[p^2]))\; R[p^2]}{}}

@@

("app"~(("app"~("prjR"~j~h))[("id", a)]))[("id", "lam" ("Wrec"~h[p^1]~b))]

@@

\begin{align*}
  \text{We first define a new type }& \goodType{\Gamma}{\Sigr{\Gamma}{n}}{}  \text{ when } \goodCtx{\Gamma}{}  \text{ is well-formed and } n \in \mathbb{N} \\
  \denotesT{\goodSig{\Gamma}{\_}{n}},\ &({\Sigr{\Gamma}{n}}), \text{ and } \ \denotesT{\cC} \text{ are mutually recursive, } \\
  & \text{defined by induction on the signature length} \\  
  \denotesT{\goodSig{\Gamma}{\_}{n}} &= (\goodTerm{\Gamma}{\_}{\Sigr{\Gamma}{n}}) \text{, and thus } \denotesT{\goodSig{\Gamma}{\sigma}{n}} \text{defined} \iff \goodTerm{\Gamma}{\denotesT{\sigma}}{\Sigr{\Gamma}{n}} \\ 
  {\Sigr{\Gamma}{(n+1)}} &= 
    \Sigma~(\Sigr{\Gamma}{n})
          ~(\Sigma~\cU
                  ~(\Sigma~(\Pi(\cC ("var"_0[p^1]))(El~"var"_0[p^1]))~\cU)) \\
  \Sigr{\Gamma}{0} &= \top \\
  \denotesT{\goodType{\Gamma}{\cC \sigma}{}}, \denotesT{\goodType{\Gamma}{\cL\sigma}{}} &\text{ defines upon } \denotesT{\sigma} 
  \text{ and inductively on the signature length} \\
  \denotesT{\goodType{\Gamma}{\cC \sigma}{}} &= \denotesT{\goodType{\Gamma}{\cL\sigma}{}} = \top \quad
      \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{0}} \\ 
  \denotesT{\goodType{\Gamma}{\cC \sigma}{}} &= 
    \Sigma~\denotesT{\cC ("pjl"~\sigma)}~(El \ ("app"~("pjr"^3~\sigma))[(p^1, "app"~("pjl"~("pjr"^2~t)))]) \\
      &\text{given } \denotesT{\goodSig{\Gamma}{\sigma}{n+1}} \\
  \denotesT{\goodType{\Gamma}{\cL\sigma}{}} &=
  \denotesT{\cL("pjl"~\sigma)} \times \Pi(El~("pjl"~("pjr"~\sigma)))(El~("app"~("pjr"^3~\sigma))) \\
  \denotesT{\goodSig{\Gamma}{\nu^+~\sigma~f~T}{n+1}} &= (\sigma, "c"~A, "lam"~f, "lam"~("c"~T)) \\ 
  \denotesT{\goodTerm{\Gamma}{\mu^+~m~t}{\cL(\nu^+~\sigma~T)}} &= (m, "lam"~t) \\ 
  \denotesT{\goodTerm{\Gamma}{\cCt m}{\cC\sigma}} & \text{ is defined upon } \denotesT{m} \text{ and inductively on the signature length} \\ 
  \denotesT{\goodTerm{\Gamma}{\cCt m}{\cC\sigma}} &= 
  (\denotesT{\cCt o}, t["sf"~f][("id", \denotesT{\cCt o})]) \\
  & \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{n+1}}, \text{ where } o = "pjl"~m, t = "app"~("pjr"~m), f = p_f\nu~\sigma \\ 
  \denotesT{\goodTerm{\Gamma}{\cCt m}{\cC \sigma}} &= () \quad \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{0}}
\end{align*}
\begin{align*}
  \denotesT{\goodInh{\Gamma}{\_}{\sigma_1}{\sigma_2}} &= \goodTerm{\Gamma, \cL\sigma_1}{\_}{\cL\sigma_2[p^1]} \\ \text{ and thus } \denotesT{\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}} &\iff  \goodTerm{\Gamma, \cL\sigma_1}{h}{\cL\sigma_2[p^1]} \\
  \denotesT{\goodInh{\Gamma}{"inhinh"~h~T~\uparrow^s}{(\nu^+ \  \sigma_1 \  T)}{(\nu^+ \  \sigma_2 \  T[(p^1, \uparrow^s)])}} &= \mu^+~(h[(p^1,p_1\mu~"var"_0)])\\ &\quad \quad \{f_2[p^1]\}~(p_2\mu'~"var"_0)[(p^1, \uparrow^s[{p^1}^{\uparrow}])] \\
\end{align*}
